\pkg{casebase}: An Alternative Framework for Survival Analysis and Comparison of Event Rates

### 2022-11-21 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
pdf compilation works after updating rmarkdown and rstudio to latest ver

Figure \ref{fig:plot-erspc-data} depicts the process of creating a population-time plot. It is built sequentially by first adding a layer for the area representing the population time in gray (Figure \ref{fig:plot-erspc-data} A), with subjects having the least amount of observation time plotted at the top of the y-axis. We immediately notice a distinctive \emph{stepwise shape} in the population time area. This is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of year from 1996 to 1999. Coupled with the uniform December 31 2006 censoring date, this led to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time (i.e.~from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and the January 1 entries in successive years. Tracked forwards in time (i.e.~from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes). Since the Swedish and Belgian centres were the last to complete recruitment in December 2003, the minimum potential follow-up is three years. Tracked further forwards in time (i.e.~after year 3) the attrition is a combination of deaths and staggered entries. As we can see, population-time plots summarise a wealth of information about the study into a simple graph.

Next, layers for the case series and base series are added. The y-axis location of each case moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area (Figure \ref{fig:plot-erspc-data} B). By randomly distributing the cases, we can get a sense of the incidence density. In Figure \ref{fig:plot-erspc-data} C, we see that more events are observed at later follow-up times. Therefore, a constant hazard model would not be appropriate in this instance as it would overestimate the incidence earlier on in time, and underestimate it later on. Finally, the base series is sampled uniformly from the study base (Figure \ref{fig:plot-erspc-data} D). The reader should refer to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population-time plot.

@@

\hypertarget{case-study-1flexible-modeling-of-the-hazard-function}{%
\subsection{Case study 1---Flexible modeling of the hazard function}\label{case-study-1flexible-modeling-of-the-hazard-function}}

@@

Modeling the hazard function directly allows us to easily visualize it with
the \texttt{plot} method and \texttt{type="hazard"} for objects of class \texttt{singleEventCB}.
We plot the hazard functions for both treatment arms in Figure \ref{fig:cs1hazard}.
The pattern we see is consistent with the population-time plot shown in Figure \ref{fig:plot-erspc-data} C, where more events are observed at later follow-up times.
The drop at the end can be explained by the fact that very few observations were followed for the entire 15 year period.

\begin{verbatim}
plot(fit, type = "hazard",
     hazard.params = list(xvar = "Follow.Up.Time",
                          by = "ScrArm"))
#> Conditions used in construction of plot
#> ScrArm: Control group / Screening group
#> offset: 0
\end{verbatim}

\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./cs1hazard-1} \caption{Estimated hazard functions for control and screening groups in the ERSPC dataset. Hazards are estimated from fitting a parametric model with casebase sampling as a function of the interaction between a cubic pspline basis (df=2) of follow-up time and treatment arm. The package vignettes provide a detailed description of how to plot hazard functions for any combination of covariates along with confidence bands.}\label{fig:cs1hazard}
\end{figure}

@@

\begin{verbatim}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14,0.1)
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
\end{verbatim}

@@

\begin{verbatim}
conf_ints <- confint(risk, fit)
head(conf_ints)
#>   time estimate conf.low conf.high      cov_prof
#> 1  0.0  0.0e+00  0.0e+00   0.0e+00 Control group
#> 2  0.1  1.8e-06  1.9e-07   1.2e-05 Control group
#> 3  0.2  3.9e-06  4.8e-07   2.2e-05 Control group
#> 4  0.3  6.1e-06  9.0e-07   3.2e-05 Control group
#> 5  0.4  8.7e-06  1.5e-06   4.1e-05 Control group
#> 6  0.5  1.2e-05  2.3e-06   5.0e-05 Control group
\end{verbatim}

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cif-conf-1} \caption{Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands.}\label{fig:erspc-cif-conf}
\end{figure}

\hypertarget{case-study-2competing-risk-analysis}{%
\subsection{Case study 2---Competing risk analysis}\label{case-study-2competing-risk-analysis}}

@@

First, we look at a population-time plot to visualize the incidence density of both relapse and the competing events. In Figure \ref{fig:compPop}, failure times are highlighted on the plot using red dots for the event of interest (panel A) and blue dots for competing events (panel B). In both panels, we see evidence of a non-constant hazard function: the density of points is larger at the beginning of follow-up than at the end.

\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./compPop-1} \caption{Population-time plot for the stem-cell transplant study with both relapse and competing events.}\label{fig:compPop}
\end{figure}

@@

\begin{verbatim}
model_cb <- fitSmoothHazard(
  Status ~ ftime + Sex + D + Phase + Source + Age,
  data = bmtcrr,
  ratio = 100,
  time = "ftime"
)
\end{verbatim}

@@

\begin{verbatim}
library(survival)
# Prepare data for coxph
bmtcrr_cox <- transform(bmtcrr, 
                        id = seq_len(nrow(bmtcrr)),
                        Status = factor(Status))

model_cox <- coxph(Surv(ftime, Status) ~ Sex + D + Phase + Source + Age,
                   data = bmtcrr_cox, id = id)
\end{verbatim}

@@

\begin{verbatim}
library(timereg)
model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
                        const(Phase) + const(Source) + const(Age),
                      data = bmtcrr, cause = 1, model = "fg")

# Estimate CI curve
risk_fg <- predict(model_fg, newdata, times = time_points)
\end{verbatim}

@@

\begin{verbatim}
# Estimate absolute risk curve
risk_cox <- survfit(model_cox, newdata = newdata)
\end{verbatim}

\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./compAbsrisk-1} \caption{Cumulative Incidence curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Fine-Gray and Aalen-Johansen estimates.}\label{fig:compAbsrisk}
\end{figure}

@@

\hypertarget{case-study-3variable-selection}{%
\subsection{Case study 3---Variable selection}\label{case-study-3variable-selection}}

@@

We then compare the risk estimation over the test set. The probabilities over time for each observation are averaged, resulting in the absolute risk curves shown in Figure \ref{fig:cs3FinalBrier} (A). We can see some minimal differences between the three models, with the Kaplan-Meier giving the lowest estimates across follow-up-time. Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard. As described above, we compare the performance between the models by computing the Brier scores over time. In Figure \ref{fig:cs3FinalBrier} (B), we can see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models. On the other hand, the differences between these three models are minimal.

\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./cs3FinalBrier-1} \caption{Comparison of Cox regression (Cox), penalized Cox regression (Pen. Cox), penalized case-base sampling estimation (Pen. CB), and Kaplan-Meier (K-M). (A)  Probability of death as a function of follow-up time. (B) Brier score as a function of follow-up time, where a lower score corresponds to better performace.}\label{fig:cs3FinalBrier}
\end{figure}

>>>>

Figure \ref{fig:plot-erspc-data} depicts the process of creating a population-time plot. It is built sequentially by first adding a layer for the area representing the population time in gray (Figure \ref{fig:plot-erspc-data}A), with subjects having the least amount of observation time plotted at the top of the y-axis. We immediately notice a distinctive \emph{stepwise shape} in the population time area. This is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of year from 1996 to 1999. Coupled with the uniform December 31 2006 censoring date, this led to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time (i.e.~from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and the January 1 entries in successive years. Tracked forwards in time (i.e.~from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes). Since the Swedish and Belgian centres were the last to complete recruitment in December 2003, the minimum potential follow-up is three years. Tracked further forwards in time (i.e.~after year 3) the attrition is a combination of deaths and staggered entries. As we can see, population-time plots summarise a wealth of information about the study into a simple graph.

Next, layers for the case series and base series are added. The y-axis location of each case moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area (Figure \ref{fig:plot-erspc-data}B). By randomly distributing the cases, we can get a sense of the incidence density. In Figure \ref{fig:plot-erspc-data}C, we see that more events are observed at later follow-up times. Therefore, a constant hazard model would not be appropriate in this instance as it would overestimate the incidence earlier on in time, and underestimate it later on. Finally, the base series is sampled uniformly from the study base (Figure \ref{fig:plot-erspc-data}D). The reader should refer to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population-time plot.

@@

\hypertarget{case-study-1flexible-modeling-of-the-hazard-function}{%
\subsection{Case study 1---flexible modeling of the hazard function}\label{case-study-1flexible-modeling-of-the-hazard-function}}

@@

Modeling the hazard function directly allows us to easily visualize it with
the \texttt{plot} method and \texttt{type="hazard"} for objects of class \texttt{singleEventCB}.
We plot the hazard functions for both treatment arms in Figure \ref{fig:cs1hazard}.
The pattern we see is consistent with the population-time plot shown in Figure \ref{fig:plot-erspc-data}C, where more events are observed at later follow-up times.
The drop at the end can be explained by the fact that very few observations were followed for the entire 15 year period.

\begin{Schunk}
\begin{Sinput}
plot(fit, type = "hazard",
     hazard.params = list(xvar = "Follow.Up.Time",
                          by = "ScrArm"))
\end{Sinput}
\begin{Soutput}
#> Conditions used in construction of plot
#> ScrArm: Control group / Screening group
#> offset: 0
\end{Soutput}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./cs1hazard-1} \caption[Estimated hazard functions for control and screening groups in the ERSPC dataset]{Estimated hazard functions for control and screening groups in the ERSPC dataset. Hazards are estimated from fitting a parametric model with casebase sampling as a function of the interaction between a cubic pspline basis (df=2) of follow-up time and treatment arm. The package vignettes provide a detailed description of how to plot hazard functions for any combination of covariates along with confidence bands.}\label{fig:cs1hazard}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14,0.1)
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
conf_ints <- confint(risk, fit)
head(conf_ints)
\end{Sinput}
\begin{Soutput}
#>   time estimate conf.low conf.high      cov_prof
#> 1  0.0  0.0e+00  0.0e+00   0.0e+00 Control group
#> 2  0.1  1.8e-06  1.9e-07   1.2e-05 Control group
#> 3  0.2  3.9e-06  4.8e-07   2.2e-05 Control group
#> 4  0.3  6.1e-06  9.0e-07   3.2e-05 Control group
#> 5  0.4  8.7e-06  1.5e-06   4.1e-05 Control group
#> 6  0.5  1.2e-05  2.3e-06   5.0e-05 Control group
\end{Soutput}
\end{Schunk}

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cif-conf-1} \caption{Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands. The smooth curve (case-base sampling) vs. step function (Cox model) highlight one of the main differences between the two approaches. The larger confidence bands in the later years is due to the relatively few number of individuals who were followed for more than 12 years.}\label{fig:erspc-cif-conf}
\end{figure}

\hypertarget{case-study-2competing-risk-analysis}{%
\subsection{Case study 2---competing risk analysis}\label{case-study-2competing-risk-analysis}}

@@

First, we look at a population-time plot to visualize the incidence density of both relapse and the competing events. In Figure \ref{fig:compPop}, failure times are highlighted on the plot using red dots for the event of interest and blue dots for competing events. In this plot, we see evidence of a non-constant hazard function: the density of points is larger at the beginning of follow-up than at the end.

\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./compPop-1} \caption{Population-time plot for the stem-cell transplant study with both relapse and competing events. The area representing the population time is shown in gray, with subjects having the least amount of observation time plotted at the top of the y-axis. The y-axis location of each case series and competing event moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area. The density of points at the beginning of follow-up relative to the end indicates a non-constant hazard function.}\label{fig:compPop}
\end{figure}

@@

\begin{Schunk}
\begin{Sinput}
model_cb <- fitSmoothHazard(
  Status ~ ftime + Sex + D + Phase + Source + Age,
  data = bmtcrr,
  ratio = 100,
  time = "ftime"
)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
library(survival)
# Prepare data for coxph
bmtcrr_cox <- transform(bmtcrr, 
                        id = seq_len(nrow(bmtcrr)),
                        Status = factor(Status))

model_cox <- coxph(Surv(ftime, Status) ~ Sex + D + Phase + Source + Age,
                   data = bmtcrr_cox, id = id)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
library(timereg)
model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
                        const(Phase) + const(Source) + const(Age),
                      data = bmtcrr, cause = 1, model = "fg")

# Estimate CI curve
risk_fg <- predict(model_fg, newdata, times = time_points)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
# Estimate absolute risk curve
risk_cox <- survfit(model_cox, newdata = newdata)
\end{Sinput}
\end{Schunk}

\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./compAbsrisk-1} \caption{Cumulative Incidence curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Fine-Gray and Aalen-Johansen estimates. In general, the three approaches agree quite well for AML, while there seems to be a difference of about 5\% between the Fine-Gray curve and the curves estimated using case-base sampling and Cox regression for ALL. However, this difference does not appear to be significant as the curve from case-base sampling is contained within a 95\% confidence band around the Fine-Gray absolute risk curve (figure not shown).}\label{fig:compAbsrisk}
\end{figure}

@@

\hypertarget{case-study-3variable-selection}{%
\subsection{Case study 3---variable selection}\label{case-study-3variable-selection}}

@@

We then compare the risk estimation over the test set. The probabilities over time for each observation are averaged, resulting in the absolute risk curves shown in Figure \ref{fig:cs3FinalBrier}A. We can see some minimal differences between the four models, with the Cox model giving the lowest estimates across follow-up-time. Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard. As described above, we compare the performance between the models by computing the Brier scores over time. In Figure \ref{fig:cs3FinalBrier}B, we can see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models. On the other hand, the differences between the three adjusted models are minimal.

\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./cs3FinalBrier-1} \caption{Comparison of Cox regression (Cox), penalized Cox regression (Pen. Cox), penalized case-base sampling estimation (Pen. CB), and Kaplan-Meier (K-M). (A)  Probability of death as a function of follow-up time. (B) Brier score as a function of follow-up time, where a lower score corresponds to better performace. The unadjusted Kaplan-Meier gives the lowest risk estimates across follow-up-time and highest Brier scores, while we can see minimal differences between the three adjusted models across both metrics. Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard.}\label{fig:cs3FinalBrier}
\end{figure}
